PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2023-01-19 main.tex
Final edits before submission

\begin{abstract}
Formation models in which terrestrial bodies grow via the pairwise
accretion of planetesimals have been reasonably successful at
reproducing the general properties of the solar system, including
small body populations. However, planetesimal accretion has not yet
been fully explored in the context of more exotic terrestrial systems,
particularly those that host short-period planets. In this work, we
use direct N-body simulations to explore and understand the growth of
planetary embryos from planetesimals in disks extending down to
$\simeq$ 1 day orbital periods. We show that planetesimal accretion
becomes nearly 100 percent efficient at short orbital periods, leading
to embryo masses that are roughly twice as large as the classical
isolation mass. For rocky bodies, the physical size of the object
begins to occupy a significant fraction of its Hill sphere at orbital
periods less than about 50 days. In this regime, most close encounters
result in collisions, rather than scattering, and the system cannot
bifurcate into a collection of dynamically hot planetesimals and
dynamically cold oligarchs, like is seen in previous work. The highly
efficient accretion seen at short orbital periods implies that systems
of tightly-packed inner planets should be almost completely devoid of
any residual small bodies. We demonstrate the robustness of our
results to assumptions about the initial disk model, and also
investigate how far material can radially mix across the boundary
between modes.
\end{abstract}

@@

Planetesimal accretion is a key phase in the terrestrial planet growth
process, bridging the gap from kilometer-sized bodies up to roughly
moon-sized objects known as planetary embryos. In the earliest stages
of the planet formation process, aerodynamic forces dominate the
growth and evolution of the solids and statistical models
\citep{johansen14, birnstiel16} are appropriate to describe how these
numerous, small bodies coagulate. Due to the internal pressure support
of the gas disk, the gas itself orbits at sub-Keplerian speed and
exerts a headwind on any solids large enough to decouple from the gas
\citep{weidenschilling77}. Above about a meter in size, this headwind
is maximally effective at sapping away angular momentum, and planet-building material can fall onto the central star on catastrophically short timescales \citep{weidenschilling77, nakagawa86}. Additionally, laboratory experiments suggest that collisions between mm- to cm- sized solids tend to result in bouncing or destruction, rather than continued growth \citep{blum93, beitz11, colwell03}. For these reasons, a number of mechanisms which involve radially concentrating solids in a planet-forming disk have been proposed to facilitate fast growth from mm to km sizes \citep{johansen07, lyra08, bai10}. Interestingly, formation models for the short-period multiplanet systems revealed by Kepler \citep{fabrycky14} also seem to require enhanced concentrations of planet-building material to reproduce the observed architectures \citep{raymond07, hansen12}.

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag plays a less significant role beyond this size. During this phase, collision cross sections are enhanced as gravitational focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because larger bodies are more effective at focusing the trajectories of nearby planetesimals, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09}. Eventually, the largest bodies (known as oligarchs) dynamically heat the surrounding planetesimals, severely limiting further growth \citep{kokubo98}. The end result of this phase is a bimodal population of dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest that the asteroid belt, Kuiper belt and the Oort cloud are largely composed of the leftovers of this stage of planet formation.

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergoes a large-scale instability
\citep{chambers98}.
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets
through a series of extremely energetic, giant 
impacts \citep{kokubo02, raymond05, raymond06}.

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received the most attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, falls into an awkward in-between, where there are too
many particles to directly track with traditional N-body codes, while
the gravitational influence of the few, but massive oligarchs that
form make statistical methods an inappropriate approach. Because of
this limitation, planetesimal accretion is usually modeled in a narrow
ring, and the results are then scaled to suit whatever situation is being studied. N-body simulations of terrestrial planet formation typically begin with a series of neatly spaced oligarchs, whose mass varies smoothly with orbital period. As we will show in this paper, this extrapolation with orbital period is inappropriate, particularly at short ($<$ 10 day) periods.

Given that systems of tightly-packed inner planets (STIPs) appear to
be a common outcome of planet formation, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical ``isolation
mass'' initial conditions used in studies of terrestrial planet formation are actually appropriate for understanding STIPs. In a series of follow up papers, we plan to use these results to directly track the formation of full-sized planets from planetesimals and to understand how gravitational perturbations from a nearby gas giant might affect the outcome of this process.

In section \ref{sec:theory} we provide an overview of the theory
behind planetesimal accretion and show that assumptions used to derive
the well-known modes of growth are only valid at sufficiently long
orbital periods. We then motivate the need for N-body simulations to
study this problem and describe the code used, along with how our
initial conditions were constructed in section \ref{sec:methods}. In
section \ref{sec:narrow}, we present a parameter study of planetesimal
accretion using a series of simulations of narrow annuli at various
orbital periods. In section \ref{sec:fulldisk} we present a set of simulations starting with a much wider planetesimal disk and demonstrate that a transition between accretion modes occurs at moderately small ($\simeq$ 50d) orbital periods. Next, we assess the impact of simplifications made to our collision model on this result in section \ref{sec:assump}. In section \ref{sec:discuss}, we discuss the implications of this multimodal accretion behavior throughout the disk for planet formation models and conclude.

@@

We begin our analysis by considering a disk of equal planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $2v/\Omega$. $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to the rms velocity, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{1}{M}\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}

In the case where the collision cross section, depends only
% TRQ: perhaps "sub-linearly"?  I see dMdt ~ M^{2/3}
on the physical size of the planetesimals, the growth scales linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \Gamma_{geo} \left( 1 + v_{esc}^2 / v^2 \right)$ \citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario, in which growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Encounters between
planetesimals that do not result in a collision play a crucial role
over long timescales. These close encounters act to convert energy
from Keplerian shear into random motion. Over time, this gravitational
stirring effect will raise the typical encounter velocities between
bodies and diminish the effectiveness of gravitational focusing. With
a spectrum of masses, these velocity differences become even more
pronounced because as the system evolves, it moves to a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 10$ for a planetesimal disk. Despite
the fact that the behavior of gravitational stirring is well-described
by the two-body formalism, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the main pathway for energy exchange between planetesimals is unable to operate.

As mentioned above, the system will tend toward a state of energy
equipartition for a non-uniform mass distribution. \citet{kokubo98}
showed that runaway growth is actually self-limiting. As the runaway
bodies grow larger, they become more effective at heating the
remaining planetesimals, which diminishes the effectiveness of
gravitational focusing and throttles the growth rate. Around the time
that the mass of the runaway bodies exceeds the mass of the planetesimals
by a factor of $\sim 50-100$ (depending on the orbital period)
\citep{ida93} a phase of less vigorous ``oligarchic'' 
growth commences, in which the few largest bodies continue to 
accrete planetesimals at similar rates.

The picture described above relies upon an important assumption, which
is that the mass distribution evolves slowly enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial planet forming region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show the ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is described
in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

In the formalism described above, the mass and velocity distribution
of the bodies are both a function of time. Due to the interdependence
of these quantities, it is not clear whether the timescales for
gravitational scattering and growth will remain proportional as the
oligarchs develop. In the case of many studies of planetesimal
accretion in the solar system
%TRQ: citations needed
{\bf(cite examples)}, the $t_{relax} \ll t_{coll}$ condition must remain true, otherwise runaway growth would have continued until all of the planetesimals were consumed. However, it isn't immediately clear what will happen if the system begins in a state where $t_{coll} \ll t_{relax}$.

An insight into the expected behavior in this regime can be gained by
defining the dimensionless parameter, $\alpha$, which is the ratio
between the physical size of a body and its Hill radius, $r_{h}$:
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its
bulk density. Assuming a fixed bulk density as bodies collide and
grow, and that no large-scale migration occurs, the scaling of both
$r_{pl}$ and $r_{h}$ with $m_{pl}^{1/3}$ means that $\alpha$ will be
constant with time. For a composition of ice and rock, $\alpha$ is
small for any populated region of the solar system ($\alpha \sim
10^{-2}$ for Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As
one moves close to the star $\alpha$ becomes larger than 1, which
implies that the physical extent of a body exceeds its Hill sphere.
{\bf(also mention that hill sphere size independent of velocity for small eccentricity)}

The size of $\alpha$ controls the relative importance of gravitational scattering and collisions in driving the evolution of the planetesimal disk. In the case that $\alpha$ is small, most close encounters will result in a gravitational interaction only, moving the system toward a state of relaxation. If, however, the Hill sphere is largely filled by the body itself, these same encounters will instead drive evolution of the masses. Because $\alpha$ is independent of time, the innermost region of the planetesimal disk, where collisions dominate over scattering events, should remain that way.

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body. Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1.

These scaling considerations motivate the range of parameters in the
following numerical experiments as we explore the different regimes.

@@

We use the tree-based N-body code {\sc ChaNGa}\footnote{A public version of {\sc ChaNGa} can be downloaded from \url{http://www-hpcc.astro.washington.edu/tools/ChaNGa.html}} to model the gravitational and collisional evolution of planetesimals at short orbital periods (where $\alpha$ is large). {\sc ChaNGa} is written using the {\sc CHARM++} parallel programming language and has been shown to perform well on up to half a million processors \citep{menon15} and can follow the evolution of gravitationally interacting swarms of up to billions of particles. Using a modified Barnes-Hut tree with hexadecapole expansions of the moments to approximate forces, {\sc ChaNGa} integrates the equations of motion using a kick-drift-kick leapfrog scheme. For all of the simulations presented in this paper, we use a node opening criteria of $\theta_{BH}$ = 0.7. Additional information about the code is available in \citep{jetley08,menon15}.

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \citet{richardson94} and \citet{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius
$2 \Delta T v + 2 r_{pl}$, where $\Delta T$ is the current timestep size
for the particle and $v$ is magnitude of its heliocentric velocity,
 are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two children. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\citep{wallace19}.
Particles are advanced in time on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

Further insight regarding the difference between the short vs long period accretion modes can be gained by looking at the 
merger history of the planetary embryos. Because all collisions are directly resolved by the N-body code, a direct lineage can be 
traced between the resulting planetary embryo and the initial planetesimals.

>>>>

\begin{abstract}
Formation models in which terrestrial bodies grow via the pairwise
accretion of planetesimals have been reasonably successful at
reproducing the general properties of the solar system, including
small body populations. However, planetesimal accretion has not yet
been fully explored in the context of more exotic terrestrial systems,
particularly those that host short-period planets. In this work, we
use direct N-body simulations to explore and understand the growth of
planetary embryos from planetesimals in disks extending down to
$\simeq$ 1 day orbital periods. We show that planetesimal accretion
becomes nearly 100 percent efficient at short orbital periods, leading
to embryo masses that are much larger than the classical
isolation mass. For rocky bodies, the physical size of the object
begins to occupy a significant fraction of its Hill sphere towards the inner edge of the disk. In this regime, most close encounters
result in collisions, rather than scattering, and the system does not
develop a bimodal population dynamically hot planetesimals and
dynamically cold oligarchs, like is seen in previous studies. The highly
efficient accretion seen at short orbital periods implies that systems
of tightly-packed inner planets should be almost completely devoid of
any residual small bodies. We demonstrate the robustness of our
results to assumptions about the initial disk model, and also
investigate that our simplified collision model has on the emergence
of this non-oligarchic growth mode in a planet forming disk.
\end{abstract}

@@

Planetesimal accretion is a key phase in the terrestrial planet growth
process, bridging the gap from kilometer-sized bodies up to roughly
moon-sized objects known as planetary embryos. In the earliest stages
of the planet formation process, aerodynamic forces dominate the
growth and evolution of the solids and statistical models
\citep{johansen14, birnstiel16} are appropriate to describe how these
numerous, small bodies coagulate.

Due to the internal pressure support
of the gas disk, the gas itself orbits at sub-Keplerian speed and
exerts a headwind on any solids large enough to decouple from the gas
\citep{weidenschilling77}. Around a meter in size, this headwind
is maximally effective at sapping away orbital angular momentum, and planet-building material can fall onto the central star on 
catastrophically short timescales \citep{weidenschilling77, nakagawa86}. Additionally, laboratory experiments suggest that 
collisions between mm- to cm- sized solids tend to result in bounces or destruction, rather than continued growth
\citep{blum93, colwell03, beitz11}.

For these reasons, a number of mechanisms to radially concentrate solids in a planet-forming disk have been proposed to 
facilitate fast growth from mm to km sizes \citep{johansen07, lyra08, bai10} in order to surmount these barriers. Interestingly, 
formation models for the short-period multiplanet systems revealed by Kepler \citep{fabrycky14} also seem to require enhanced 
concentrations of planet-building material to reproduce the observed architectures \citep{raymond07, hansen12}.

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag 
plays an increasingly unimportant role beyond this size. During this phase, collision cross sections are enhanced as gravitational 
focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because gravitational focusing 
becomes more effective as bodies grow larger, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09} and a 
power law spectrum of masses develop. Eventually, the largest bodies (known as oligarchs) dynamically heat the remaining 
planetesimals, severely limiting further growth \citep{kokubo98}. The final outcome of this phase is a bimodal population of 
dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest 
that the asteroid belt, Kuiper belt and the Oort cloud are largely composed of the leftovers of this stage of planet formation.

After a long period of quiescence, the collection of embryos and
remaining planetesimals undergo a large-scale instability
\citep{chambers98}.
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets
through a series of extremely energetic giant 
impacts \citep{kokubo02, raymond05, raymond06}.

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received most of the attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, is more difficult and expensive to model because there are too
many particles to directly track with traditional N-body codes, while
the gravitational influence of the few massive oligarchs that
form make statistical methods inappropriate. Due to computational expense, N-body simulations of
planetesimal accretion are usually modeled in a narrow
ring \citep{kokubo96, kokubo98}, and the results and timescales are then scaled to suit whatever relevant orbital period is being 
studied. N-body simulations of terrestrial planet assembly typically begin with a series of neatly spaced oligarchs, whose mass 
varies smoothly with orbital period. As we will show in this paper, oligarchic growth does not scale to arbitrarily short
orbital periods.

Given that systems of tightly-packed inner planets (STIPs) appear to
be a common outcome of planet formation \citep{lantham11, lissauer11, rowe14}, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical ``isolation
mass'' initial conditions used in studies of terrestrial planet formation are actually appropriate for understanding STIPs. In a series of follow up papers, we plan to use these results to directly track the formation of full-sized planets from planetesimals and to understand how gravitational perturbations from a nearby gas giant might affect the outcome of this process.

In section \ref{sec:theory} we provide an overview of the theory
behind planetesimal accretion and show that assumptions used to derive
the well-known modes of growth are only valid at sufficiently long
orbital periods. We then motivate the need for N-body simulations to
study this problem and describe the code used, along with how our
initial conditions were constructed in section \ref{sec:methods}. In
section \ref{sec:narrow}, we present a parameter study of planetesimal
accretion using a series of simulations of narrow annuli that exhibit both oligarchic and non-oligarchic growth. In section 
\ref{sec:fulldisk} we present a set of simulations starting with a  wide planetesimal disk and demonstrate that a transition 
between accretion modes occurs at moderately small orbital periods. Next, we assess the impact of 
simplifications made to our collision model on this result in section \ref{sec:assump}. In section \ref{sec:discuss}, we discuss the 
implications of this multimodal accretion behavior throughout the disk for planet formation models and conclude.

@@

We begin our analysis by considering a disk of equal mass planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $2v/\Omega$). $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to the rms velocity, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}
Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{1}{M}\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}

In the case where the collision cross section depends only
on the physical size of the planetesimals, the growth scales sub-linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, $\Gamma = \pi r_{pl}^2 \left( 1 + v_{esc}^2 / v^2 \right)$ \citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario where growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Gravitational stirring, which converts
Keplerian shear into random motion, raises the typical encounter velocity
between planetesimals over time \citep{weidenschilling89, ida90} and diminishes the effectiveness of gravitational
focusing. As the mass spectrum of the system evolves away from uniformity, these velocity differences become even more
pronounced. As the system evolves, it tends toward a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random 
motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body 
relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 10$ for a planetesimal disk. Despite
the fact that the behavior of gravitational stirring is well-described
by a two-body formula, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters, which 
requires a three-body formalism. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the 
Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the 
main pathway for energy exchange between planetesimals and growing protoplanets is unable to operate.

As mentioned above, the system will tend toward a state of energy
equipartition for a non-uniform mass distribution. \citet{kokubo98}
showed that runaway growth is actually self-limiting. As the runaway
bodies develop, they become increasingly effective at heating the
remaining planetesimals, which diminishes the
gravitational focusing cross sections and throttles the growth rate. Around the time
that the mass of the runaway bodies exceeds the mass of the planetesimals
by a factor of $\sim 50-100$
\citep{ida93} a phase of less vigorous ``oligarchic'' 
growth commences, in which the largest bodies continue to 
accrete planetesimals at similar rates, independent of mass.

The picture described above relies upon an crucial assumption, which
is that the mass distribution evolves slowly enough for gravitational stirring to maintain energy equipartition. In other words, the relaxation timescale must remain short relative to the growth timescale. For typical conditions near the terrestrial region of the solar system, this timescale condition is satisfied. Due to the steep dependence of the relaxation time on encounter velocity, however, this condition can easily be violated at shorter orbital periods.

In figure \ref{fig:timescales}, we show the ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The eccentricity dispersion is 
scaled by the Hill factor $\left( m_{pl}/ 3 M_{\star}
\right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary
between shear and dispersion dominated encounters. The horizontal
dashed line indicates where $t_{relax} = t_{coll}$. The timescale
criterion for oligarchic growth is only satisfied in regions where the
disk is sufficiently dynamically cold and the orbital period is
sufficiently long. In sections \ref{sec:narrow} and \ref{sec:fulldisk}
we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

In the formalism described above, the mass and velocity distribution
of the bodies are both a function of time. Due to the interdependence
of these quantities, it is not clear whether the timescales for
gravitational scattering and growth will remain proportional as the
oligarchs develop. In the case of the solar system, $t_{relax} \ll t_{coll}$ likely continued to remain true, otherwise runaway 
growth would have consumed all of the small bodies and there would be nothing left to populate the asteroid or Kuiper belt. In 
the case where $t_{coll} \ll t_{relax}$, however, it is not clear how the system might evolve.

An insight into the expected behavior in this regime can be gained by
defining a dimensionless parameter, $\alpha$, which is the ratio
between the physical size of a body and its Hill radius, $r_{h}$
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its
bulk density. Assuming that the bulk density stays constant as bodies collide and
grow, and that no large-scale migration occurs, the scaling of both
$r_{pl}$ and $r_{h}$ as $m_{pl}^{1/3}$ means that $\alpha$ will be
constant with time. For a composition of ice and rock, $\alpha$ is
small for any presently populated region of the solar system ($\alpha \sim
10^{-2}$ near Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As
one moves closer to the sun, $\alpha$ becomes larger than 1, which
implies that the Hill sphere of a body becomes smaller than its physical size.

The magnitude of $\alpha$ controls the relative importance of gravitational scattering and collisions in driving the evolution of the planetesimal disk. When $\alpha$ is small, most close encounters will result in a gravitational interaction, moving the system toward a state of relaxation. If, however, the Hill sphere is largely filled by the body itself, these same encounters will instead drive evolution of the mass spectrum. Because $\alpha$ stays constant with mass, the boundary in the disk where collisions or gravitational encounters dominate, will stay static with time.

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body. Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1 and the collision cross section is mostly set by the geometric value.

These scaling considerations motivate the range of parameters we choose for the numerical experiments presented in the next section, where we aim to understand where and when runaway and oligarchic growth can operate.

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/alpha_beta_mass.png}
    \caption{The final state of the mass distributions for the
      narrow annulus simulations described in section \ref{sec:narrow}. For small
      $\alpha$, a few embryos form alongside a power law tail of
      planetesimals. For larger values of $\alpha$, the mass distribution stops being bimodal. 
      As in the previous figure, the initial choice of $\beta$ does not appear to have any meaningful impact on the end result.
      \label{fig:alpha_beta_mass}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/alpha_beta_evo.png}
    \caption{The evolution of the ratio between the maximum and mean mass for the four simulations presented in section 
    \ref{sec:narrow}. The runaway growth phase can be identified by a positive trend in this ratio. For all values of $\alpha$, an 
    increase in $\beta$ has the effect of delaying runaway growth.\label{fig:alpha_beta_evo}}
\end{center}
\end{figure}

@@

We use the tree-based N-body code {\sc ChaNGa}\footnote{A public version of {\sc ChaNGa} can be downloaded from \url{http://www-hpcc.astro.washington.edu/tools/ChaNGa.html}} to model the gravitational and collisional evolution of a swarm of planetesimals. {\sc ChaNGa} is written using the {\sc CHARM++} parallel programming language and has been shown to perform well on up to half a million processors \citep{menon15} and can follow the evolution of gravitationally interacting collections of up to billions of particles. Using a modified Barnes-Hut tree with hexadecapole expansions of the moments to approximate forces, {\sc ChaNGa} integrates the equations of motion using a kick-drift-kick leapfrog scheme. For all of the simulations presented in this paper, we use a node opening criteria of $\theta_{BH}$ = 0.7. Additional information about the code is available in \citep{jetley08,menon15}.

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \citet{richardson94} and \citet{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius
$2 \Delta T v + 2 r_{pl}$, where $\Delta T$ is the current timestep size
for the particle and $v$ is magnitude of its heliocentric velocity,
 are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two children. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\citep{wallace19}.
Particles are advanced on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

Further insight regarding the difference between the short vs long period accretion modes can be gained by looking at the 
growth history of the planetary embryos. Because all collisions are directly resolved by the N-body code, a lineage can be 
traced between each planetary embryo and the initial planetesimals.
